---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# aRtpalette

The goal of `aRtpalette` is to make it easy to bring art-inspired color palettes into R-based data visualizations. The package lets you store palettes as structured objects, preview them as color swatches, and expand them when a chart needs more colors than the source artwork provides. 

The inspiration for this package came from the frustration of having to use the colors from paintings in my ggplot2 charts, but there was no clean workflow for doing so in R. As I used R, I had to manually copy hex codes and hope the colors looked right once applied. I wanted a tool that treated color as data, something you could store in a table, validate, visualize, and manipulate programmatically. 

## Installation

You can install the development version of `aRtpalette` from
[GitHub](https://github.com/) with:

``` r
# Option 1: using devtools
# install.packages("devtools")
devtools::install_github("ADC-405-S26/aRtpalette")

# Option 2: using pak
# install.packages("pak")
pak::pak("ADC-405-S26/aRtpalette")
```
## The built-in dataset

The package includes `artwork_palettes`, a table of 20 color palettes extracted from iconic artworks across art history. Each row is one painting, with five hex color codes and metadata including the artist, artwork title, and art movement.

```{r}
library(aRtpalette)
library(ggplot2)
data(artwork_palettes)
head(artwork_palettes[, c("palette_name", "artist", "style", "hex1", "hex2", "hex3")])
```

## Core functions

### `palette_from_hex()` — store colors as a validated object

```{r}
row <- artwork_palettes[artwork_palettes$palette_name == "starry_night", ]

starry_night <- palette_from_hex(
  hex_codes = c(row$hex1, row$hex2, row$hex3, row$hex4, row$hex5),
  name      = "starry_night"
)

starry_night
```

The function validates every hex code on input and returns a clear error message if any is malformed. So problems surface immediately rather than silently breaking a chart later. 

### `plot_palette()` — preview before you apply

```{r README-plot-palette, fig.height = 2}
plot_palette(starry_night)
```

Seeing the palette as swatches before applying it to real data saves the back-and-forth of guessing whether the colors will work together. 

### `apply_palette()` — expand for any chart size

```{r}
# Exactly 3 colors for a 3-group chart
apply_palette(starry_night, n = 3)
```

```{r}
# Expand to 8 colors via interpolation
apply_palette(starry_night, n = 8)
```

If you ask for more colors than the palette contains, the function interpolates between the existing colors, allowing you to take the original color and expand it creatively for any chart size. 

## A full workflow example

Here we pull a palette directly from `artwork_palettes`, preview it as swatches, then apply it to a chart, showing all three functions working together in one clean pipeline.
```{r}

```{r README-step1, fig.height = 2}
# Step 1 — pull the Starry Night row from our dataset
# and turn it into a validated palette object
row <- artwork_palettes[artwork_palettes$palette_name == "starry_night", ]

pal <- palette_from_hex(
  c(row$hex1, row$hex2, row$hex3, row$hex4, row$hex5),
  name = "starry_night"
)

# Step 2 — preview the palette before using it
# so we know exactly what colors we are working with
plot_palette(pal)
```

The swatch view lets us confirm the colors look right before committing them to a chart. This will prevent wasting time going back and forth in the later process.

```{r README-step2, fig.height = 4}
# Step 3 — count artworks by art style from our own dataset
style_counts <- as.data.frame(table(artwork_palettes$style))
colnames(style_counts) <- c("style", "count")

# Step 4 — expand the palette to exactly the number of 
# style groups we need using apply_palette()
chart_colors <- apply_palette(pal, n = nrow(style_counts))

# Step 5 — plug the colors directly into ggplot2
ggplot(style_counts, aes(x = reorder(style, count),
                          y = count,
                          fill = style)) +
  geom_bar(stat = "identity", width = 0.7) +
  scale_fill_manual(values = chart_colors) +
  coord_flip() +
  labs(
    title    = "Number of Artworks by Art Style",
    subtitle = "Colors from Van Gogh's The Starry Night",
    x        = "Art Style",
    y        = "Count"
  ) +
  theme_classic() +
  theme(legend.position = "none")
```

The palette came from `artwork_palettes`, was previewed with `plot_palette()`, expanded to fit the number of style groups with `apply_palette()`, and applied directly to the chart. No hex codes were typed by hand.

## Summary

`aRtpalette` provides three functions that work together as a pipeline:

1.**`palette_from_hex()`** — store hex codes from any artwork as a validated, named palette object

2.**`plot_palette()`** — preview the palette as color swatches before applying it to a chart

3.**`apply_palette()`** — extract or expand the palette to exactly the number of colors your chart needs.

Combined with the built-in `artwork_palettes` dataset of 20 palettes from iconic artworks, the package makes it easy to bring art history into R without any manual hex code lookup.


## Learn more

See the full vignette for a detailed walkthrough including palette
comparison across art movements, the complete art analysis workflow,
and more ggplot2 examples:

```r
vignette("getting-started-with-aRtpalette")
```
